# Reproducible Research: Peer Assessment 1


## Loading and preprocessing the data

```{r, echo=TRUE}
library(knitr)
library(ggplot2)
library(reshape2)
library(xtable)

unzip("activity.zip")
activity <- read.csv("activity.csv", colClasses=c("numeric", "Date", "numeric"))
act.na.rm <- activity[!is.na(activity$steps),]

#tick markers and their labels. Used for plots
days.at <- seq(min(activity$date), max(activity$date), by=7)
days.at.int <- as.numeric(days.at) - as.numeric(min(days.at))
days.lab <- as.character(days.at, format="%b %d")

int.at <- seq(0,2400, by=300)
int.lab <- as.character(strptime(sprintf("%04d", int.at), format="%H%M"),
                        format="%I%p")
```

## What is mean total number of steps taken per day?

Here is the histogram of total number of steps taken per day
```{r, echo=TRUE,fig.height=10,fig.width=10}
act.sum <- tapply(act.na.rm$steps, act.na.rm$date, sum)

#labels for this barplot
days.at.1 <- as.Date(labels(act.sum)[[1]], format="%Y-%m-%d")
days.lab.1 <- as.character(days.at.1, format="%b %d")

barplot(act.sum, 
        names.arg=days.lab.1, axis.lty=1,
        col=grey(seq(0.9, 0.1, length.out=7)),
        xlab="Day", ylab="Total steps",
        main="Total number of steps taken per day")
```

Here is of mean and median of total number of steps taken per day
```{r, echo=TRUE,fig.height=10,fig.width=10}
act.mean <- mean(act.sum)
act.median <- median(act.sum)
act.stats <- rbind(mean=act.mean, median=act.median)
colnames(act.stats) <- c("value")

act.stats
```

## What is the average daily activity pattern?
```{r, echo=TRUE,fig.height=10,fig.width=10}
act.pat <- tapply(act.na.rm$steps, act.na.rm$interval, mean)

act.pat.melt <- melt(act.pat)
names(act.pat.melt) <- c("interval", "steps.mean")

#getting the max point and creating its label
max.pt <- order(act.pat.melt$steps.mean, decreasing=TRUE)[1]
max.pt <- act.pat.melt[max.pt,]
max.pt.lab <- "label"
max.pt.lab <- paste(round(max.pt$steps.mean, 2), "@", 
                    as.character(strptime(sprintf("%04d", max.pt$interval), 
                                          format="%H%M"),
                                 format="%I:%M%p"))
                    
plot(act.pat.melt$interval, act.pat.melt$steps.mean, type="l", xaxt="n",
     xlab="Time", ylab="Average number of steps",
     main="Average daily activity pattern")
axis(1, at=int.at, labels=int.lab, cex.axis=0.8)
points(max.pt$interval, max.pt$steps.mean, col="red", pch=16)
text(max.pt$interval, max.pt$steps.mean, max.pt.lab, pos=4, col="red")
```

## Imputing missing values

There are **`r sum(is.na(activity))` NA values** in activity.csv. Imputing those with the mean of interval column  

Here is the histogram of imputed total number of steps taken per day

```{r, echo=TRUE,fig.height=10,fig.width=10}
act.na <- activity[is.na(activity),]

#reusing act.pat.melt from previous block
act.na.imp <- merge(act.na, act.pat.melt, by.x="interval", by.y="interval")
act.imp <- act.na.imp[, c("steps.mean","date","interval")]

names(act.imp) <- c("steps", "date", "interval")
act.imp <- rbind(act.na.rm, act.imp)

#getting the sum for each day
act.imp.sum <- tapply(act.imp$steps, act.imp$date, sum)

#labels for this barplot
days.at.1 <- seq(min(act.imp$date), max(act.imp$date), by=1)
days.lab.1 <- as.character(days.at.1, format="%b %d")

barplot(act.imp.sum, 
        names.arg=days.lab.1, axis.lty=1,
        col=grey(seq(0.9, 0.1, length.out=7)),
        xlab="Day", ylab="Total steps",
        main="Total number of steps taken per day (imputed)")
```

Here are the tables of mean and median of imputed total number of steps taken per day

```{r, echo=TRUE,fig.height=10,fig.width=10}
act.imp.mean <- mean(act.imp.sum)
act.imp.median <- median(act.imp.sum)
act.imp.stats <- rbind(mean=act.imp.mean, median=act.imp.median)
colnames(act.imp.stats) <- c("value")

act.imp.stats
```


## Are there differences in activity patterns between weekdays and weekends?
```{r, echo=TRUE,fig.height=10,fig.width=10}

#adding at new factor variable to indicate if the day is weekday or weekend
act.imp$day <- weekdays(as.Date(act.imp$date, format="%Y-%m-%d"), TRUE)
act.imp$weekend <- act.imp$day %in% c("Sat","Sun")
act.imp$day.type <- factor(act.imp$weekend, labels=c("Weekday", "Weekend"))

#plot the graph with facets for weekday and weekend
qp.mean <- qplot(interval, steps, data=act.imp, facets=day.type ~ ., 
                stat="summary", fun.y="mean", geom="line",
                ylab="Average steps per 5 minute interval", xlab="Time",
                main="Activity pattern comparision of weekday and weekend") + 
                scale_x_continuous(breaks = int.at,labels = int.lab)
qp.mean
```
